### 2016-05-12 responses/response.tex

\noindent
Contextual dependencies can often be satisfied in multiple ways\footnote{For example, see the
discussion about distributed programming in Section~3.2.2 of the thesis.}. By capturing the
contextual properties in the type-and-coeffect system, we are making the existing
run-time ambiguities \emph{more explicit}. Thus the proposal is to focus on systems with
\emph{coeffect-directed semantics} rather than modifying the calculi that inspired the work
in the thesis in order to force coherence.

\section{Resolving ambiguity}
\label{sec:ambiguity}
\vspace{-1em}

The case where the semantics of a context-aware program depends on the typing derivation is best
demonstrated using the system for implicit parameters (Section~3.2.1 of the thesis). Consider the
following example:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{f} = (~\kvd{let}~\ident{?x} = 10~\kvd{in}~
  \lambda \ident{v} \rightarrow \ident{?x})~\kvd{in}\\[-0.2em]
\kvd{let}~\ident{?x} = 15~\kvd{in}~\ident{f}~0
\end{array}
\end{equation*}
%
Here, a function value $\lambda \ident{v} \rightarrow \ident{?x}$ is created in a context where the
implicit parameter $\ident{?x}$ has a value 10. The function is then called in a context where
$\ident{?x}$ has a value 15. Is the result of the expression $10$ or $15$? The choice of the
run-time semantics needs to be reflected in the type system.

\paragraph{Resolution in Haskell.}
In Haskell, the implicit parameters from the \emph{declaration-site} are preferred and so the result
is 10. Correspondingly, the type system only permits one typing -- with the parameter \ident{?x}
required on the declaration-side (allowing no rebinding at the call site). However, if the function
value was defined in a context not containing \ident{?x}, the value from the call site would be used.
Thus, Haskell uses a \emph{defaulting mechanism} for implicit parameters, similar to that of F\#
operators discussed in Section~\ref{sec:background-practice}.

@@

This claim is left implicit and we intend to address this in the final version 
(by proving that the three concrete examples spelled out in Chapter 3 do, indeed, fit
the common patterns identified later in Chapters 4 and 5).

\vspace{-0.5em}
\paragraph{4.5. Universal type systems.}

The thesis does not clearly explain the benefit of introducing the universal type systems
based on the coeffect calculi, as we do not prove the usual type-safety property for the concrete systems.
This becomes clearer when we consider the system as \emph{type erasure}.

The type systems guarantee that \emph{coeffect indices} will match in the resulting 
translation. The relationship from indices to values is injective in the general case, but 
we can also show a number of important properties (which will be clarified in the revised
version) for the concrete systems, \eg~an upper bound on the number of required
past values in causal dataflow.

@@

% ==================================================================================================

\section{Summary}

An interesting aspect of many context-aware computations is that they require access to some
context, but do not explicitly specify where the context comes from. In case of implicit parameters,
the values can be provided either by the call site or by the declaration-site.

### 2016-05-12 main.tex
Context-aware systems edits (Chapter 3)
Edits in implementation, consistency (Chapter 7)

  We start with a detailed discussion of the motivation for the thesis and an overview of our
  methodology (Chapter~\ref{ch:intro}). Next, we discuss previous programming langauge research that
  leads to the work presented in this thesis (Chapter~\ref{ch:pathways}) and we examine a number
  of practical context-aware systems in detail (Chapter~\ref{ch:applications}), identifying two
  kinds of context that we later capture by \emph{flat} and \emph{structural coeffects}.
}
\part{Context-aware programming}
\label{part:context-aware}

\include{text/Chapter01}
\include{text/Chapter02}
\include{text/Chapter03}

@@

\ctparttext{In this part, we capture the similarities between the concrete context-aware langauges
  presented in the previous chapter. We also develop the key novel technical contributions of
  the thesis. We define a \emph{flat coeffect type system} (Chapter~\ref{ch:flat}) that is
  parameterized by a \emph{coeffect algebra} and a mechanism for choosing unique typing derivation.
  We instantiate a coeffect type system with a concrete coeffect algebra and procedure for choosing
  unique typing derivation for three languages to capture dataflow, implicit parameters and liveness.

@@

\ctparttext{ TBD }
\part{Towards practical coeffects}
\label{part:context-aware}

%\include{text/Chapter07}
%\include{text/Chapter08}
%\include{text/Chapter09}

>>>>

  We start with a detailed discussion of the motivation for the thesis and an overview of our
  methodology (Chapter~\ref{ch:intro}). Next, we discuss previous programming language research that
  leads to the work presented in this thesis (Chapter~\ref{ch:pathways}) and we examine a number
  of practical context-aware systems in detail (Chapter~\ref{ch:applications}), identifying two
  kinds of context that we later capture by \emph{flat} and \emph{structural coeffects}.
}
\part{Context-aware programming}
\label{part:context-aware}

%\include{text/Chapter01}
%\include{text/Chapter02}
%\include{text/Chapter03}

@@

\ctparttext{In this part, we capture the similarities between the concrete context-aware languages
  presented in the previous chapter. We also develop the key novel technical contributions of
  the thesis. We define a \emph{flat coeffect type system} (Chapter~\ref{ch:flat}) that is
  parameterized by a \emph{coeffect algebra} and a mechanism for choosing unique typing derivation.
  We instantiate a coeffect type system with a concrete coeffect algebra and procedure for choosing
  unique typing derivation for three languages to capture dataflow, implicit parameters and liveness.

@@

\ctparttext{ In the first part of the thesis, we argued for the importance of \emph{context} in
  programming languages. As programs execute in increasingly diverse and rich environments,
  languages need to understand and check how programs use such context. In the second part, we
  developed theoretical foundations (type system and semantics) for context-aware programming
  languages. What remains to be done if context-aware programming languages are to become
  ``the next big thing''?

  In this part, we explore practical aspects of implementing context-aware programming languages
  based on coeffects and related future work. We discuss a prototype implementation (Chapter~\ref{ch:impl}),
  which links together all parts of the theory discussed in the previous part. Building a
  production-ready programming language is outside the scope of the thesis, so we instead focus
  on conveying the concept of coeffects to broader audience and make the implementation available
  as a web-based interactive essay (Section~\ref{sec:impl-essay}) at: \url{http://tomasp.net/coeffects}.

  In further work (Chapter~\ref{ch:further}), we outline unification of flat and structural
  coeffects (Section~\ref{sec:further-unified}), which may be more suited for embedding in practical
  languages and we discuss alternative approaches for using coeffects in programming languages
  (\ref{sec:further-meta}).
}

\part{Towards practical coeffects}
\label{part:context-aware}

\include{text/Chapter07}
\include{text/Chapter08}
\include{text/Chapter09}

### 2016-05-12 text/Abstract.tex
Edits in implementation, consistency (Chapter 7)

We identify a number of program properties that were not connected before, but model some notion 
of context. Our examples include tracking different execution platforms (and their versions) 
in cross-platform development, resources available in different execution environments (\eg~GPS
sensor on a phone and database on the server), but also more traditional notions such as 
variable usage (\eg~in liveness analaysis and linear logics) or past values in 
stream-based data-flow programming.

>>>>

We identify a number of program properties that were not connected before, but model some notion 
of context. Our examples include tracking different execution platforms (and their versions) 
in cross-platform development, resources available in different execution environments (\eg~GPS
sensor on a phone and database on the server), but also more traditional notions such as 
variable usage (\eg~in liveness analaysis and linear logics) or past values in 
stream-based dataflow programming.

### 2016-05-12 text/AppendixA.tex
Edits in implementation, consistency (Chapter 7)

This appendix gives full typing derivations for the examples presented in Chapter~\ref{ch:applications},
which demonstrate simple coeffect calculi for implicit parameters, liveness and data-flow.

@@

\section{Coeffect typing for data-flow}
\label{sec:appendixa-dataflow}

The Section~\ref{sec:applications-flat-dataflow} presents a coeffect type system that tracks the 
maximal number of past values required by a data-flow computation. The discussion includes the 
following example:
%
\begin{equation*}
\begin{array}{l}
(\,   \kvd{if}~~(\kvd{prev}~\ident{tick})=0\\[-0.25em]
\,\,\, \kvd{then}~(\lambda x \rightarrow \kvd{prev}~x)\\[-0.25em]
\,\,\, \kvd{else}~(\lambda x \rightarrow x) \,)\qquad(\kvd{prev}~\ident{counter})
\end{array}
\end{equation*}
%
In order to give typing for the above example, we first need to extend the language with 
conditionals. The typing rule for the $\kvd{if}$ construct is:
%
\begin{equation*}
\tyrule{if}
  { \coctx{\Gamma}{\cclrd{m}} \vdash e : \ident{bool} &
    \coctx{\Gamma}{\cclrd{n}} \vdash e_1 : \tau &
    \coctx{\Gamma}{\cclrd{n}} \vdash e_2 : \tau }
  {\coctx{\Gamma}{\textnormal{max}(\cclrd{m},\cclrd{n})} \vdash \kvd{if}~e~\kvd{then}~e_1~\kvd{else}~e_2 : \tau_1 }
\end{equation*}
%
Given a condition that requires $\cclrd{m}$ past and two (alternative) bodies that each require
$\cclrd{n}$ past values, the composed expression requires at most the maximum of the two, \ie~the
same context is passed to both the condition and the body using point-wise composition. As we will
see, the fact that both branches require the same number of past values means that we need to use
the subcoeffecting rule.

@@

\figcaption{Coeffect typing for data-flow}
\label{fig:appendixa-dfl}
\end{figure}

>>>>

This appendix gives full typing derivations for the examples presented in Chapter~\ref{ch:applications},
which demonstrate simple coeffect calculi for implicit parameters, liveness and dataflow.

@@

\section{Coeffect typing for dataflow}
\label{sec:appendixa-dataflow}

The Section~\ref{sec:applications-flat-dataflow} presents a coeffect type system that tracks the 
maximal number of past values required by a dataflow computation. The discussion includes the 
following example:
%
\begin{equation*}
\begin{array}{l}
(\,   \kvd{if}~~(\kvd{prev}~\ident{tick})=0\\[-0.25em]
\,\,\, \kvd{then}~(\lambda x \rightarrow \kvd{prev}~x)\\[-0.25em]
\,\,\, \kvd{else}~(\lambda x \rightarrow x) \,)\qquad(\kvd{prev}~\ident{counter})
\end{array}
\end{equation*}
%
In order to give typing for the above example, we first need to extend the language with 
conditionals. The typing rule for the $\kvd{if}$ construct is:
%
\begin{equation*}
\tyrule{if}
  { \coctx{\Gamma}{\cclrd{m}} \vdash e : \ident{bool} &
    \coctx{\Gamma}{\cclrd{n}} \vdash e_1 : \tau &
    \coctx{\Gamma}{\cclrd{n}} \vdash e_2 : \tau }
  {\coctx{\Gamma}{\textnormal{max}(\cclrd{m},\cclrd{n})} \vdash \kvd{if}~e~\kvd{then}~e_1~\kvd{else}~e_2 : \tau_1 }
\end{equation*}
%
Given a condition that requires $\cclrd{m}$ past and two (alternative) bodies that each require
$\cclrd{n}$ past values, the composed expression requires at most the maximum of the two, \ie~the
same context is passed to both the condition and the body using point-wise composition. As we will
see, the fact that both branches require the same number of past values means that we need to use
the subcoeffecting rule.

@@

\figcaption{Coeffect typing for dataflow}
\label{fig:appendixa-dfl}
\end{figure}

### 2016-05-12 text/Chapter01.tex
Edits in implementation, consistency (Chapter 7)

The final example leaves the topic of cross-platform and distributed computing. We focus on
checking how arrays are accessed. This is a simpler version of the data-flow programming
examples used later in the thesis.

@@

\item In Chapter~\ref{ch:applications}, we present the first contribution of the thesis -- the discovery
  of the connection between a number of existing programming language features that are related
  to context. The chapter presents type systems and semantics for a number of systems and analyses
  (including data-flow, liveness analysis, distributed programming and Haskell's type classes).
  Our novel presentation reveals their similarity.

>>>>

The final example leaves the topic of cross-platform and distributed computing. We focus on
checking how arrays are accessed. This is a simpler version of the dataflow programming
examples used later in the thesis.

@@

\item In Chapter~\ref{ch:applications}, we present the first contribution of the thesis -- the discovery
  of the connection between a number of existing programming language features that are related
  to context. The chapter presents type systems and semantics for a number of systems and analyses
  (including dataflow, liveness analysis, distributed programming and Haskell's type classes).
  Our novel presentation reveals their similarity.

### 2016-05-12 text/Chapter02.tex
Context-aware systems edits (Chapter 3)
Edits in implementation, consistency (Chapter 7)

\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{f}=\\[-0.25em]
\qquad\kvd{let}~\ident{x}=10~\kvd{in}\\[-0.25em]
\qquad\lambda \ident{y}\rightarrow \ident{x}+\ident{y}~\kvd{in}\\[-0.25em]
\kvd{let}~\ident{x}=5~\kvd{in}\\
\ident{f}~0
\end{array}
\end{equation*}
%
The program can be evaluated in two ways, depending on the variable binding mechanism:
%
\begin{itemize}
\item \textsc{Static (lexical) binding.} In a langauge with static binding (such as ML or Java),
  the variable \ident{x} inside the body of the lambda function is statically bound to the
  declaration in the lexical scope -- that is, the variable on the second line -- and the
  expression evaluates to 10.

@@

\noindent
When we view variable access as a context requirement, we can see that the body of the function
($\ident{x}+\ident{y}$) requires a context that provides values for variables \ident{x} and
\ident{y}. In static binding, the context requirements of the body can be placed on the scope in
which the function is defined (declaration-site). In dynamic binding, the requirements are
\emph{delayed} and are placed on the scope in which the function is called (call-site).

@@

\subsection{Implicit parameter binding}

@@

The following two examples are variations on the one discussed in the Section~\ref{sec:path-binding-var}.
We replace a variable \ident{x} with an implicit parameter \ident{?x}. On the left, the
implicit parameter is defined both at declaration-site and at the call-site. On the right, the
implicit parameter is available only at the call-site:

\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{f}=\\[-0.25em]
\qquad\kvd{let}~\ident{?x}=10~\kvd{in}\\[-0.25em]
\qquad\lambda \ident{y}\rightarrow \ident{?x}+\ident{y}~\kvd{in}\\[-0.25em]
\kvd{let}~\ident{?x}=5~\kvd{in}\\
\ident{f}~0
\end{array}\hspace{5em}\begin{array}{l}
\kvd{let}~\ident{f}=\\[-0.25em]
\qquad\lambda \ident{y}\rightarrow \ident{?x}+\ident{y}~\kvd{in}\\[-0.25em]
\kvd{let}~\ident{?x}=5~\kvd{in}\\
\ident{f}~0
\end{array}
\end{equation*}
%
The binding rules for implicit parameters can be summarized as ``\emph{static binding when possible,
dynamic binding when needed}''. If an implicit parameter is available in scope, then the value
is statically bound and the context requirement is satisfied using the declaration-site context.
Otherwise, the context requirement is delayed and has to be satisfied at the call-site.
In the example on the left, \ident{?x} is bound to $10$ the function \ident{f} has no
delayed context requirements and the expression evalautes to $10$. In the example on the right,
the context requirement \ident{?x} is delayed and is satisfied via dynamic binding when calling
the function. The expression evaluates to $5$.

@@

In this thesis, we consider different notions of context, using implicit parameters as just one
of several motivating examples. However, implicit parameters are a valuable example, because they
clearly illustrate the ambiguity inherent in context-aware programs -- the context requirements of
a function can be satisfied using the context available at declaration-site or using the context
available at the call-site.

We also aim to find a description of context-aware languages that does not make ad-hoc decisions
about how context requirements are split between the declaration-site and the call-site. While
Haskell's solution for implicit parameters might be the most reasonable one, different notions
of context might require different domain-specific choices and the general framework of
context-aware programming should make that possible.

@@

\item The compiler chooses an appropriate typing derivation. In the above case, this is done
  based on the context in which the expression appears, but other options are possible
  (in some cases, there is a \emph{default} resolution; in some cases, the compiler requires
  explicit typing annotation).
\end{itemize}
%
The function $\lambda\ident{y}\rightarrow\ident{?x}+\ident{y}$ in Haskell also has multiple
possible typing derivations and its semantics varies depending on the type. If the lexical
scope contains a binding for $\ident{?x}$, the function type is $\kvd{int}\rightarrow\kvd{int}$
and it captures the value from the lexical scope. Otherwise, the type of the function is
$\{\ident{?x}:\kvd{int}\}\Rightarrow\kvd{int}\rightarrow\kvd{int}$ and it reads the parameter
value from a hidden dictionary that is passed together with the input from the call-site.

@@

The theoretical parts of this thesis extend the work of Uustalu and Vene who use comonads
to give the semantics of data-flow computations \cite{comonads-dataflow} and more generally,
notions of \emph{context-dependent computations} \cite{comonads-notions}. The computations discussed
in the latter work include streams, arrays and containers. This is a more diverse set of examples,
but they all mostly represent forms of collections. Ahman et al. \cite{comonads-containers} discuss
the relation between comonads and \emph{containers} \cite{types-containers} in more details.

@@

In causal data-flow, we can interpret the list as a list consisting of past values, with the
current value in the head. Then, the $\ident{cobind}$ operation calculates the current value
of the output based on the current and all past values of the input; the second element is
calculated based on all past values and the last element is calculated based just on the initial
input $(l_n)$. In addition to the operations of comonad, the model also uses some operations that
are specific to causal data-flow:
%
\begin{equation*}
\begin{array}{rcl}
\ident{prev}~(l_1, \ldots, l_n) &=& (l_2, \ldots, l_n)\\
\end{array}
\end{equation*}
%
The operation drops the first element from the list. In the data-flow interpretation, this means
that it returns the previous state of a value.

Now, consider a simple data-flow language with single-variable contexts, variables,
primitive built-in functions and a construct $\kvd{prev}~e$ that returns the previous
value of the computation $e$. We omit the typing rules, but they are simple -- assuming $e$
has a type $\tau$, the expression $\kvd{prev}~e$ has also type $\tau$. The fact that
the language models data-flow and values are lists (of past values) is a matter of semantics,
which is defined as follows:
%
\begin{equation*}
\begin{array}{rcl}
\sem{x\!:\!\tau \vdash x : \tau} &=& \ident{counit}_\ctyp{}{}\\
\sem{x\!:\!\tau_1 \vdash f~e : \tau_3} &=& f \circ (\ident{cobind}_\ctyp{}{} ~\sem{e})\\
\sem{x\!:\!\tau_1 \vdash \kvd{prev}~e : \tau_2} &=& \ident{prev} \circ (\ident{cobind}_\ctyp{}{} ~\sem{e})\\
\end{array}
\end{equation*}
%
The semantics follows that of effectful computations using monads. A variable access is interpreted
using $\ident{counit}_\ctyp{}{}$ (extract the variable value); composition
uses $\ident{cobind}_\ctyp{}{}$ to propagate the context to the function $f$ and $\kvd{prev}$
is interpreted using the primitive $\ident{prev}$ (which takes a list and returns a list).
For example, the judgement $x\!:\!\tau \vdash \kvd{prev}~(\kvd{prev}~x) : \tau$ represents an
expression that expects context with variable $x$ and returns a stream of values before the
previous one. The semantics of the term expresses this behaviour:
$(\ident{prev} \circ \ident{prev} \circ (\ident{cobind}_\ctyp{}{}~\ident{counit}_\ctyp{}{}))$.
Note that the first operation is simply an identity function thanks to the comonad laws discussed
earlier.

>>>>

\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{f}=\\[-0.25em]
\qquad\kvd{let}~\ident{x}=10~\kvd{in}\\[-0.25em]
\qquad\lambda \ident{y}\rightarrow \ident{x}+\ident{y}~\kvd{in}\\[-0.25em]
\kvd{let}~\ident{x}=5~\kvd{in}\\
\ident{f}~0
\end{array}
\end{equation*}
%
The program can be evaluated in two ways, depending on the variable binding mechanism:
%
\begin{itemize}
\item \textsc{Static (lexical) binding.} In a language with static binding (such as ML or Java),
  the variable \ident{x} inside the body of the lambda function is statically bound to the
  declaration in the lexical scope -- that is, the variable on the second line -- and the
  expression evaluates to 10.

@@

\noindent
When we view variable access as a context requirement, we can see that the body of the function
($\ident{x}+\ident{y}$) requires a context that provides values for variables \ident{x} and
\ident{y}. In static binding, the context requirements of the body can be placed on the scope in
which the function is defined (declaration-site). In dynamic binding, the requirements are
\emph{delayed} and are placed on the scope in which the function is called (call site).

@@

\subsection{Implicit parameter binding}
\label{sec:path-binding-impl}

@@

The following two examples are variations on the one discussed in the Section~\ref{sec:path-binding-var}.
We replace a variable \ident{x} with an implicit parameter \ident{?x}. On the left, the
implicit parameter is defined both at declaration-site and at the call site. On the right, the
implicit parameter is available only at the call site:

\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{f}=\\[-0.25em]
\qquad\kvd{let}~\ident{?x}=10~\kvd{in}\\[-0.25em]
\qquad\lambda \ident{y}\rightarrow \ident{?x}+\ident{y}~\kvd{in}\\[-0.25em]
\kvd{let}~\ident{?x}=5~\kvd{in}\\
\ident{f}~0
\end{array}\hspace{5em}\begin{array}{l}
\kvd{let}~\ident{f}=\\[-0.25em]
\qquad\lambda \ident{y}\rightarrow \ident{?x}+\ident{y}~\kvd{in}\\[-0.25em]
\kvd{let}~\ident{?x}=5~\kvd{in}\\
\ident{f}~0
\end{array}
\end{equation*}
%
The binding rules for implicit parameters can be summarized as ``\emph{static binding when possible,
dynamic binding when needed}''. If an implicit parameter is available in scope, then the value
is statically bound and the context requirement is satisfied using the declaration-site context.
Otherwise, the context requirement is delayed and has to be satisfied at the call site.
In the example on the left, \ident{?x} is bound to $10$ the function \ident{f} has no
delayed context requirements and the expression evalautes to $10$. In the example on the right,
the context requirement \ident{?x} is delayed and is satisfied via dynamic binding when calling
the function. The expression evaluates to $5$.

@@

In this thesis, we consider different notions of context, using implicit parameters as just one
of several motivating examples. However, implicit parameters are a valuable example, because they
clearly illustrate the ambiguity inherent in context-aware programs -- the context requirements of
a function can be satisfied using the context available at declaration-site or using the context
available at the call site.

We also aim to find a description of context-aware languages that does not make ad-hoc decisions
about how context requirements are split between the declaration-site and the call site. While
Haskell's solution for implicit parameters might be the most reasonable one, different notions
of context might require different domain-specific choices and the general framework of
context-aware programming should make that possible.

@@

\item The compiler chooses an appropriate typing derivation. In the above case, this is done
  based on the context in which the expression appears, but other options are possible
  (in some cases, there is a \emph{default} resolution; in some cases, the compiler requires
  explicit typing annotation).
\end{itemize}
%
The function $\lambda\ident{y}\rightarrow\ident{?x}+\ident{y}$ in Haskell also has multiple
possible typing derivations and its semantics varies depending on the type. If the lexical
scope contains a binding for $\ident{?x}$, the function type is $\kvd{int}\rightarrow\kvd{int}$
and it captures the value from the lexical scope. Otherwise, the type of the function is
$\{\ident{?x}:\kvd{int}\}\Rightarrow\kvd{int}\rightarrow\kvd{int}$ and it reads the parameter
value from a hidden dictionary that is passed together with the input from the call site.

@@

The theoretical parts of this thesis extend the work of Uustalu and Vene who use comonads
to give the semantics of dataflow computations \cite{comonads-dataflow} and more generally,
notions of \emph{context-dependent computations} \cite{comonads-notions}. The computations discussed
in the latter work include streams, arrays and containers. This is a more diverse set of examples,
but they all mostly represent forms of collections. Ahman et al. \cite{comonads-containers} discuss
the relation between comonads and \emph{containers} \cite{types-containers} in more details.

@@

In causal dataflow, we can interpret the list as a list consisting of past values, with the
current value in the head. Then, the $\ident{cobind}$ operation calculates the current value
of the output based on the current and all past values of the input; the second element is
calculated based on all past values and the last element is calculated based just on the initial
input $(l_n)$. In addition to the operations of comonad, the model also uses some operations that
are specific to causal dataflow:
%
\begin{equation*}
\begin{array}{rcl}
\ident{prev}~(l_1, \ldots, l_n) &=& (l_2, \ldots, l_n)\\
\end{array}
\end{equation*}
%
The operation drops the first element from the list. In the dataflow interpretation, this means
that it returns the previous state of a value.

Now, consider a simple dataflow language with single-variable contexts, variables,
primitive built-in functions and a construct $\kvd{prev}~e$ that returns the previous
value of the computation $e$. We omit the typing rules, but they are simple -- assuming $e$
has a type $\tau$, the expression $\kvd{prev}~e$ has also type $\tau$. The fact that
the language models dataflow and values are lists (of past values) is a matter of semantics,
which is defined as follows:
%
\begin{equation*}
\begin{array}{rcl}
\sem{x\!:\!\tau \vdash x : \tau} &=& \ident{counit}_\ctyp{}{}\\
\sem{x\!:\!\tau_1 \vdash f~e : \tau_3} &=& f \circ (\ident{cobind}_\ctyp{}{} ~\sem{e})\\
\sem{x\!:\!\tau_1 \vdash \kvd{prev}~e : \tau_2} &=& \ident{prev} \circ (\ident{cobind}_\ctyp{}{} ~\sem{e})\\
\end{array}
\end{equation*}
%
The semantics follows that of effectful computations using monads. A variable access is interpreted
using $\ident{counit}_\ctyp{}{}$ (extract the variable value); composition
uses $\ident{cobind}_\ctyp{}{}$ to propagate the context to the function $f$ and $\kvd{prev}$
is interpreted using the primitive $\ident{prev}$ (which takes a list and returns a list).
For example, the judgement $x\!:\!\tau \vdash \kvd{prev}~(\kvd{prev}~x) : \tau$ represents an
expression that expects context with variable $x$ and returns a stream of values before the
previous one. The semantics of the term expresses this behaviour:
$(\ident{prev} \circ \ident{prev} \circ (\ident{cobind}_\ctyp{}{}~\ident{counit}_\ctyp{}{}))$.
Note that the first operation is simply an identity function thanks to the comonad laws discussed
earlier.
